---
title: "R Tutorials: Using R Markdown"
author: "Chros LeBoa"
date: "8/2/2019"
output: word_document
---

## R Markdown 

The tutorial for  <http://rmarkdown.rstudio.com>.
The book on r Markdown: I should consult this for any more questions I might have 
When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

Code chunks are deliniated with ```{r} and three more of those things to close it  
```{r cars}
summary(cars)
```


### To add code chunks Press Command, Option and I 
there is also an add chunk command in the tool bar 

inside a code chunk {} can use 
  incclude = false - chunk will not be in final doc
  echo = false - code is not presented but the results are in doc 
  message = false - no messages from code in final doc 
  warning =flase - no warnings will show in final 
  fig.cap ="" will add a caption to any figures in final doc 
  
  for example

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


  
Global options apply to all code chunks 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Inline Code
  You can insert functions by using `r` followed by the function name or dataframe name -- have multiple defs for same code and then uncomment the one you want to try in the function 
  R will include results of code and apply yext but not the code 
  
  
R markdown can add code in bash and python + others 
To use another language use ```{CODE LANGUAGE NAME} insteaf of just R 
You can still use the code chunks 

#Parameters: 
They are good to use when want to do the same report with different datasets for dif regions or specific period of time 
  you can use the params: command to determine parameters to determine which dataset to use  
  use params$<parameter name> to call dataset or other parameters 
  can use render that will use a new set of parameters 
if you use knit with parameters button in dropdown menu it will let you preview the visualization with different parameters 

#Tables 
Usually in normal R formatting but can use knitr::kable to make fancier tables 
Make sure to use results = "asis" to keep knitr from doing too much formatting 

Somw foos options for making tables 
* [xtable](https://cran.r-project.org/web/packages/xtable/)
* [stargazer](https://cran.r-project.org/web/packages/stargazer/)
* [pander](http://rapporter.github.io/pander/)
* [tables](https://cran.r-project.org/web/packages/tables/)
* [ascii](http://eusebe.github.io/ascii/)
* etc.

Table code is here 
```{r echo = TRUE, results = 'asis'}
library(knitr)
kable(mtcars[1:5, ], caption = "A knitr kable.")
```

Use Pandoc Markdown to deal with headers, footers, etc 
it says to check out the options in the R markdown general guide

*italics*
**bold**
`code`
[links](https://rmarkdown.rstudio.com/lesson-8.html)

can also make equations $a=1+2$
and bibliographies []

#Output 
  use Render command or go up to the knit menu above 
  in render use output = "pdf" "word_document" or other things 
  
Document examples
html_notebook ⧉ - Interactive R Notebooks
html_document ⧉ - HTML document w/ Bootstrap CSS
pdf_document ⧉ - PDF document (via LaTeX template)
word_document ⧉ - Microsoft Word document (docx)
odt_document ⧉ - OpenDocument Text document
rtf_document ⧉ - Rich Text Format document
md_document ⧉ - Markdown document (various flavors)
Presentations (slides)
ioslides_presentation ⧉ - HTML presentation with ioslides
revealjs::revealjs_presentation ⧉ - HTML presentation with reveal.js
slidy_presentation ⧉ - HTML presentation with W3C Slidy
beamer_presentation ⧉ - PDF presentation with LaTeX Beamer
powerpoint_presentation ⧉: PowerPoint presentation
More
flexdashboard::flex_dashboard - Interactive dashboards
tufte::tufte_handout ⧉ - PDF handouts in the style of Edward Tufte
tufte::tufte_html ⧉ - HTML handouts in the style of Edward Tufte
tufte::tufte_book ⧉ - PDF books in the style of Edward Tufte
html_vignette ⧉ - R package vignette (HTML)
github_document - GitHub Flavored Markdown document

#R_Notebooks
  R markdown has a notebook type that 
  way more infor here https://bookdown.org/yihui/rmarkdown/notebook.html
  but it allows for iteration of code 
  
  ###use output:html_notebook to save as a notebook 
  if want normal console can choose chunk output in console 
  
  In morkdown chunk is sent at once but in notebook only one line at a time 
  can use fig.width, fig.heigt and fig.asp to change the size of figures in the notebook 
  Graphs appear inline unless set up viewer
  
rmd_stub = "examples/r-notebook-stub.Rmd"
cat(readLines(rmd_stub), sep = "\n")
---
title: "R Notebook Stub"
output: html_notebook
---

```{r chunk-one}
print("Hello, World!")
```
  
#Presentations
can save as ppt, or other types of slides 
use *** to make manual slide break 
use >- to make bullets 

#Dashboards 
can hve dashboards that have multiple types of information on them 
[file here](https://rstudio.cloud/project/181966)

Each # begins a new page in dashboard 
each ## begins a new column 
each ### begins new box 

can use{.sidebar} to make more chanes 
There is a whole seperate training on flexidash that i can check out later 
[flexidash](https://rmarkdown.rstudio.com/flexdashboard/using.html)

**This would be cool to report the number of studies wrong for Kristen 

each thing in a dashboard has a notes componant 
use > to add notes after it 

Within flexidash can also make storyboards with the data 
*** adds commentary next to the code you are making 

[can make websites too](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html)

#Interactive docs
Can use Java to make interactive visualizations - allows to look at maps or other things 

[example](https://rstudio.cloud/project/181983)

##Leaflet 
seems really cool and has been used for a lot of visualizations
[Leaflet Link](http://rstudio.github.io/leaflet/) It is an API that can be opened in R 
--> I will take notes on it in a seperate document 

##Shiny 

A method of making web maps using R 
use runtime:shiny to start a shiny app 

[Shiny example](https://rstudio.cloud/project/181983)

For other tutorial on Shiny use [this link](http://shiny.rstudio.com/)

#Cheatsheets 
are built into the help window of R markdown







