---
title: 'R Tutorials: Intro to R on Code Club'
author: "Chris LeBoa"
date: "8/2/2019"
output: html_notebook
---
### What this is
  I have taken a year of classes in stats that use R but given the ample time I have in Nepal I wanted to check the holes that I have in my understanding of the language and best practices in it. 

#Introduction to R 
  
  
  Do not put spaces in row names from Excel 
 
  use <- to assign variables -- you can get this  function by doing (option and -) at once 
  
  When go to a file can use (option , command, c) to get the path name instead of just the file /Users/ChrisLeBoa/Desktop/Learning R Better/IntroTutorials/data/CC-1-RBasics-master/edidiv.csv
  
```{r}
library(dplyr)

getwd() #shows the directory things will be save into/ where I am working 
edidiv <- read.csv("/Users/ChrisLeBoa/Desktop/Learning R Better/IntroTutorials/data/CC-1-RBasics-master/edidiv.csv") #File path of where the data is stored 

head(edidiv$taxonGroup)     # Displays the first few rows of this column only
class(edidiv$taxonGroup)    # Tells you what type of variable we're dealing with: it's character now but we want it to be a factor

edidiv$taxonGroup <- as.factor(edidiv$taxonGroup) #turn data categorical


```
  # Types of data 
           categorical = factor 
           Numbers = nummeric 
           Whole Numbers = integer
           Strings = character 
  
  Use $ after the df name to get a single object 
  dim () shows # of rows and columns 
  
  
  #Splitting Dataset 
  
  use filter and will seperate dataset based on groups 
  
```{r}
#split data by group
summary(edidiv$taxonGroup)
Beetle <- filter(edidiv, taxonGroup == "Beetle")
Bird <- filter(edidiv, taxonGroup == "Bird") 
Butterfly <- filter(edidiv, taxonGroup == "Butterfly") 
Dragonfly <- filter(edidiv, taxonGroup == "Dragonfly") 
Flowering.Plants <- filter(edidiv, taxonGroup == "Flowering.Plants") 
Lichen <- filter(edidiv, taxonGroup == "Lichen") 
Fungus <- filter(edidiv, taxonGroup == "Fungus") 
Liverwort <- filter(edidiv, taxonGroup == "Liverwort") 
Hymenopteran <- filter(edidiv, taxonGroup == "Hymenopteran")
Mammal <- filter(edidiv, taxonGroup == "Mammal") 
Mollusc <- filter(edidiv, taxonGroup == "Mollusc") 
```
Unique shows the number of uniwue variables 
length counts them


```{r, include=FALSE}
#Code stores each variable as unique species
a <- length(unique(Beetle$taxonName))
b <- length(unique(Bird$taxonName))
c <- length(unique(Butterfly$taxonName))
d <- length(unique(Dragonfly$taxonName))
e <- length(unique(Flowering.Plants$taxonName))
f <- length(unique(Fungus$taxonName))
g <- length(unique(Hymenopteran$taxonName))
h <- length(unique(Lichen$taxonName))
i <- length(unique(Liverwort$taxonName))
j <- length(unique(Mammal$taxonName))
k <- length(unique(Mollusc$taxonName))

```
#Plotting Data

use c and add variable names to concatenate (add them all to one list)

Once add them all to list need to make names for each - making sure to keep in the same order is important

using barplot will make a plot
can use help to get help

By using png() it will save a photo of your plot
```{r}
biodiv <- c(a,b,c,d,e,f,g,h,i,j,k)     # We are chaining together all the values; pay attention to the object names you have calculated and their order
names(biodiv) <- c("Beetle",
                   "Bird",
                   "Butterfly",
                   "Dragonfly",
                   "Flowering.Plants",
                   "Fungus",
                   "Hymenopteran",
                   "Lichen",
                   "Liverwort",
                   "Mammal",
                   "Mollusc")

barplot(biodiv)

#Save Barplot to computer
png("barplot.png", width=1600, height=600)  # look up the help for this function: you can customise the size and resolution of the image
barplot(biodiv, xlab="Taxa", ylab="Number of species", ylim=c(0,600), cex.names= 1.5, cex.axis=1.5, cex.lab=1.5)
dev.off()
# The cex code increases the font size when greater than one (and decreases it when less than one). 
```
#Dataframes 

Vectors are good at dealing with one set of values but dataframes are good for multiple types of values 

##Important to always keep copy of raw data and then save manipulations as a seperate file 

use write.csv or write_excel to do that 

use function data.frame to turn list of vectors into a dataframe.  
```{r}
# Creating an object called "taxa" that contains all the taxa names
taxa <- c("Beetle",
          "Bird",
          "Butterfly",
          "Dragonfly",
          "Flowering.Plants",
          "Fungus",
          "Hymenopteran",
          "Lichen",
          "Liverwort",
          "Mammal",
          "Mollusc")
# Turning this object into a factor, i.e. a categorical variable
taxa_f <- factor(taxa)

# Combining all the values for the number of species in an object called richness
richness <- c(a,b,c,d,e,f,g,h,i,j,k)

# Creating the data frame from the two vectors
biodata <- data.frame(taxa_f, richness)

# Saving the file
write.csv(biodata, file="biodata.csv")  # it will be saved in your working directory

barplot(biodata$richness)
```

### Challenge Section
   Save barplot of wingspan to computer
```{r}
# Calculate the mean wingspan for each bird species. The function to do that is simply: mean()
sparrow <- mean(22, 24, 21)
kingfisher <- mean(26, 23, 25)
eagle <- mean(195, 201, 185)
hummingbird <- mean(8, 9, 9)

# Chain them together in a vector
wingspan <- c(sparrow, kingfisher, eagle, hummingbird)

# Create a bird species vector (careful to match the order of the previous vector!)
bird_sp <- c("sparrow", "kingfisher", "eagle", "hummingbird")
# notice how we put quotation marks around the names. It's because we're creating (character) values; writing sparrow without the "" would call the object we created in the code above, which would return the value 22!

# Bird species is currently in character form, but it should be a factor. Let's fix that:
# (To be honest it does not make any difference to the output here, but it would for some other types of plot. Take good habits early!)
class(bird_sp)                      # currently character
bird_sp <- as.factor(bird_sp)       # transforming into factor
class(bird_sp)                      # now a factor! 


# Then, combine the two vectors in a data frame
wings <- data.frame(bird_sp, wingspan)

# Plot the bar plot & save it to file

png("wingspan_plot.png", width=800, height=600)
barplot(wings$wingspan, names.arg = wings$bird_sp,    # notice how we call the bird_sp column instead of typing all the names
        xlab = "Bird species", 
        ylab = "Average wingspan (cm)",               # adding axis titles
        ylim = c(0, 200),                             # setting the limits of the y axis to fit the eagle
        col = "gold"                                  # changing the colour because why not!
        )
dev.off()
```
 
##########################################################
#Intro to R 2: Troubleshooting R 

Common errors 
   R not being updated 
   Syntax 
   R cant recognize function 
  
  use package::function (ie dplyr::filter) to return info 
  
  if problem with for loops 
    good to use R's debugging tool 
    
   Data not in right format
      use str()
      or summary() to check types of variables 
      
      
  Save versions of scripts in case R crashes - check out git tutorial 
  
  if you press insert on keyboard code gets overwritten
  
  To fix do help with package name or google 
  
  




